---
title: "Luminex Xponent: Plate-level Quality Control"
author: "Sahal Thahir & Jeff Bailey"
date: "2024-06-04"
params:
  input_file: "data/raw/luminex/RTSS_Ahero_IgG3_3.21.24_Lowbeadtest.csv"  # Path to the input CSV file
  min_beadcount: 50  # Minimum bead count for well bead QC
  min_Rsquared: 0.90 # Minimum R-squared for passing standard curve QC
  file_path: "data/qc"  # Output filepath for plate-level plots and PDFs 
  background: "BSA" # Name the analyte column for background subtraction
  standards: ["AMA1", "CelTos", "HRPII"] # List of standard analytes, use ["Analyte1", "Analyte2", etc.]
output:
  pdf_document:
    toc: true  # Include table of contents
---

## Aim

This code aims to perform plate-level quality control analysis for Luminex
studies through bead counts, Background MFI, and Standard curve
analysis.

### Instructions

1. **Modify YAML Parameters:**
   - Open this document in RStudio.
   - Edit the YAML parameters at the top of the document:
     - `input_file`: Path to your input CSV file.
     - `min_beadcount`: Set the minimum bead count threshold.
     - `min_Rsquared`: Set the minimum R-squared value for standard curve QC.
     - `file_path`: Specify the output directory for plots and PDFs.
     - `

2. **Run the Analysis:**
   - After modifying the YAML parameters, click the *knit* button at the top of the page to generate the report.

### Code Explanation
#### Dependencies
- `r R.version.string`
  - `tidyverse` (v. `r packageVersion("tidyverse")`)
  - `here` (v. `r packageVersion("here")`)
  - `gridExtra` (v. `r packageVersion("gridExtra")`)

This code chunk will ensure these packages are installed and loaded.

```{r setup, include=FALSE}
#No input required
   knitr::opts_chunk$set(echo = TRUE)
    knitr::opts_knit$set()
    # Function to check if a package is installed and install it if it's not
    install_if_missing <- function(packages) {
      for (package in packages) {
        if (!require(package, character.only = TRUE)) {
          install.packages(package, dependencies = TRUE)
          library(package, character.only = TRUE)
        }
      }
    }
    
    # List of required packages
    required_packages <- c("tidyverse", "here", "gridExtra")
    
    # Install missing packages
    install_if_missing(required_packages)

```

#### Import and Data Extraction
1. **Raw data import:** Import raw data CSV file into R. 

```{r Plate QC- raw data import + raw data extraction, echo=TRUE, warning=TRUE}
# No input required
  #YAML extraction
     # Read parameters from YAML header
      input_file <- params$input_file
      min_beadcount <- params$min_beadcount
      min_Rsquared <- params$min_Rsquared
      file_path <- params$file_path
      background_col <- params$background
  
  # Raw data import
      lines <- readLines(input_file)
```    


2. **Extract dataframes:** find median MFI and bead count data. This includes a check to ensure the median MFI and beadcount dataframes align. 

```{r Plate QC- Extract Median MFI and bead count, echo=FALSE, fig.height=17, fig.width=10, warning=FALSE}
# No input required

# Median MFI 
    # Extracting Median MFI data from the raw file
    # Median MFI dataframe: Extract data between "Median" and two rows above "Net MFI"
    # Find the row index containing the word "Median"
    median_row_index <- grep("Median", lines)
    
    # Find the row index containing the word "Net MFI"
    net_mfi_row_index <- grep("Net MFI", lines)
    
    # Find the row index of two rows above "Net MFI"
    net_mfi_row_index <- net_mfi_row_index - 2
    
    # Extract data between "Median" and two rows above "Net MFI"
    data_lines <- lines[(median_row_index + 1):net_mfi_row_index]
    
    # Create a data frame from the extracted lines
    medianMFI_df <- read.csv(text = paste(data_lines, collapse = "\n"), na.strings = "")
    
    # Trim medianMFI_df to include only columns between "Location" and the column before "Total.Events"
    location_index <- grep("Location", colnames(medianMFI_df))
    total_events_index <- grep("Total.Events", colnames(medianMFI_df))
    medianMFI_df <- medianMFI_df[, c(location_index, (location_index + 1):(total_events_index - 1))]

# Bead count dataframe (beadcount_df)
    # Extracting Bead Count data from the raw file
    # Read lines from the input file
    lines <- readLines(input_file)
    
    # Identify the start row for bead count data
    start_row <- grep("DataType", lines)
    start_row <- start_row[grepl("Count", lines[start_row])]
    start_row <- start_row[!grepl("Per Bead", lines[start_row])]
    start_row <- max(start_row) + 1
    
    # Identify the end row for bead count data
    end_row <- grep("Avg Net MFI", lines) - 2
    
    # Extract the beadcount_df data frame
    if (length(start_row) > 0 & length(end_row) > 0) {
      beadcount_df <- read.csv(text = paste(lines[start_row:end_row], collapse = "\n"))
      
      # Trim beadcount_df to include only columns between "Location" and the column before "Total.Events"
      location_index <- grep("Location", colnames(beadcount_df))
      total_events_index <- grep("Total.Events", colnames(beadcount_df))
      beadcount_df <- beadcount_df[, c(location_index, (location_index + 1):(total_events_index - 1))]
    } else {
      beadcount_df <- NULL
    }

# Ensure data alignment
    # Ensure no missing values in the location and sample columns
      if (any(is.na(medianMFI_df$Location)) | any(is.na(beadcount_df$Location))) {
        stop("Missing values found in the 'Location' column.")
      }
      
      if (any(is.na(medianMFI_df$Sample)) | any(is.na(beadcount_df$Sample))) {
        stop("Missing values found in the 'Sample' column.")
      }
      
         # Ensure all column names match between the two data frames
      if (!identical(colnames(medianMFI_df), colnames(beadcount_df))) {
        stop("The column names do not match between the median MFI and bead count data frames.")
      }
      
      # If the column names are identical
      print("Column names are consistent between the median MFI and bead count data frames.")
```

3. **Identify low bead count wells:** this searches the beadcount data frame for well-antigen data with low bead count, as set in the YAML.This will export a csv file of the samples which will need to be rerun. 

```{r Plate QC- Low bead count list, echo=FALSE}
# No input required
    # Replace bead counts below the minimum threshold with "Failed QC (bead)"
    for (col in names(beadcount_df)[sapply(beadcount_df, is.numeric)]) {
      beadcount_df[[col]][beadcount_df[[col]] < params$min_beadcount] <- "Failed QC (bead)"
    }
    
    # Create a data frame of wells that failed QC
    beadqc_low_df <- beadcount_df %>%
      pivot_longer(cols = -c(Location, Sample), names_to = "Antigen", values_to = "Value") %>%
      filter(Value == "Failed QC (bead)") %>%
      select(Location, Sample, Antigen) %>%
      mutate(Plate = gsub("\\.csv$", "", basename(params$input_file)))
    
    # Specify the output file path
    output_file_list <- file.path(params$file_path, 
                                  paste0(tools::file_path_sans_ext(basename(params$input_file)), "_beadqc_low_df.csv"))
    
    # Export the data frame of low wells as a CSV file
    write.csv(beadqc_low_df, file = output_file_list, row.names = FALSE)
    
    # Print a message indicating the export was successful
    cat("The bead QC list has been exported to", output_file_list, "\n")
              
```
### Visualization of bead count data
This code generates a well schematic identifying wells identified as having low bead counts. as well as a heatmap of antigen/well locations and the quantity of beads, to assist in determining the cause of low bead counts (low antigen concentration in mastermix, vortexing between wells, etc. )

```{r Plate QC- Well Schematic, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
 # Create well schematic
                # Extract row and column indices from Location
                beadqc_low_df$row_col <- gsub(".*,(.*)\\)", "\\1", beadqc_low_df$Location)
                beadqc_low_df$row <- gsub("\\(.*", "", gsub(".*\\)", "", beadqc_low_df$Location))
                beadqc_low_df$col <- gsub(".*\\)", "", gsub("\\(.*", "", beadqc_low_df$Location))
                
                # Create plate layout with row_col column
                plate_layout <- expand.grid(
                  row = LETTERS[1:8],
                  col = as.character(1:12)
                )
                plate_layout$row_col <- paste0(plate_layout$row, plate_layout$col)
                
                # Merge with low well locations to mark low wells
                plate_layout$is_low <- plate_layout$row_col %in% beadqc_low_df$row_col
                
                # Generate the plot
                title <- paste("96-well schematic of Plate: ", gsub(".csv", "", basename(input_file)))
                plate_plot <- ggplot(plate_layout, aes(x = col, y = row, fill = is_low)) +
                  geom_tile(color = "black") +
                  scale_fill_manual(values = c("deepskyblue3", "darkred"), labels = c("Passed", "Low bead count")) +
                  labs(x = "Column", y = "Row", title = title, fill = "Bead Count") +
                  theme_minimal() +
                  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1),
                        axis.text.y = element_text(angle = 0, vjust = 0.5, hjust = 1)) +
                  guides(fill = guide_legend(title = "Bead count QC"))
                
                # Modify the y-axis to have the top row as "A" and the bottom row as "H"
                plate_plot <- plate_plot + scale_y_discrete(limits = rev(levels(factor(plate_layout$row))))
                
       # Display the plot
          plate_plot

        # Display list
          selected_columns <- beadqc_low_df %>% select(Location, Sample, Antigen, Plate)
        selected_columns
        
        # Print a message indicating the export was successful
        cat("The bead QC list has been exported to", output_file_list, "\n")
        
# Generate the heatmap 
        if (!is.null(beadcount_df)) {
          # Convert the dataframe to long format
          beadqc_long <- beadcount_df %>% pivot_longer(cols = -c(Location, Sample), names_to = "Numeric_Column", values_to = "Value")
          
          # Convert the "Value" column to numeric
          beadqc_long$Value <- as.numeric(beadqc_long$Value)
          
          # Create the title
          title <- paste("Quality Control: Bead count for each antigen", gsub(".csv", "", basename(input_file)))
          
          # Create the heatmap
          plot <- ggplot(beadqc_long, aes(x = Numeric_Column, y = factor(Location, levels = rev(unique(beadqc_long$Location))), fill = Value)) +
            geom_tile(color = "black") +
            scale_fill_gradient(low = "red", high = "greenyellow", limits = c(0, max(beadqc_long$Value, na.rm = TRUE)), breaks = seq(0, max(beadqc_long$Value, na.rm = TRUE), by = 50)) +
            labs(x = "Antigen", y = "Location", title = title) +  # Set the title
            theme_minimal()
          
          # Define the output file path
          output_file <- file.path(file_path, paste0(tools::file_path_sans_ext(basename(input_file)), "_beadqc.png"))
          
          # Export the plot to file_path
          ggsave(filename = output_file, plot = plot, width = 10, height = 6, dpi = 300)
          
          # Print a message indicating the export was successful
          cat("The bead QC report has been exported to", output_file, "\n")
          
          # Display the plot
          print(plot)
        }
```

4. **Modify Median MFI data:** All plate locations with a low beadcount with be switched to a `NA` value for standard curve evaluation 

```{r Plate QC- Modify MFI for low bead counts, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
# Modify the medianMFI_df to mark low bead count wells with NA
    # Identify the rows in medianMFI_df that correspond to the low bead count wells
    low_bead_locations <- beadqc_low_df$Location
    
    # Iterate over each numeric column in medianMFI_df (excluding Location and Sample columns)
    for (col in names(medianMFI_df)[sapply(medianMFI_df, is.numeric)]) {
      medianMFI_df[[col]][medianMFI_df$Location %in% low_bead_locations] <- NA
    }
    
    # Print a message indicating the modification was successful
    cat("Median MFI data has been modified for low bead count wells.\n")
```

5. **Evaluate standard curves:** Make sure the background and standard analytes are defined in the YAML params

This first chunk conducts background MFI subtraction, based on the analyte defined in the YAML. 
```{r Plate QC- background median MFI subtraction, echo=FALSE, fig.height=8, fig.width=12, warning=FALSE}
# Subtract background MFI from median MFI values
          
          # Get the name of the background MFI column
          background_column <- params$background
          
          # Check if the background column exists in medianMFI_df
          if (!background_column %in% names(medianMFI_df)) {
            stop(paste("Background column", background_column, "not found in medianMFI_df"))
          }
          
          # Extract background MFI values
          background_values <- medianMFI_df[[background_column]]
          
          # Subtract background MFI from each numeric column (excluding Location, Sample, and the background column itself)
          for (col in names(medianMFI_df)[sapply(medianMFI_df, is.numeric) & names(medianMFI_df) != background_column]) {
            medianMFI_df[[col]] <- medianMFI_df[[col]] - background_values
          }
          
          # Print a message indicating the subtraction was successful
          cat("Background MFI subtraction completed.\n")
```
This code chunk 
```{r Plate QC- plate standard plots, echo=FALSE, fig.height=14, fig.width=8, warning=FALSE}
# No input required

# Analytes: Get the names of the specified standard analytes from params
analytes <- params$standards

# Check if the specified analytes exist in the medianMFI_df
valid_analytes <- analytes[analytes %in% names(medianMFI_df)]

# Warn if any specified analytes are not found in the data frame
if (length(valid_analytes) != length(analytes)) {
  warning("The following analytes were not found in the data frame and will be ignored: ", 
          paste(setdiff(analytes, valid_analytes), collapse = ", "))
}

# Initialize plate_qc as an empty dataframe
plate_qc <- data.frame(Analyte = character(), R_squared = numeric())

# Function to calculate correlation and R-squared
calculate_correlation <- function(df) {
  lm_model <- lm(Log10_MFI ~ Dilution_Factor, data = df)
  correlation <- cor(df$Log10_MFI, df$Dilution_Factor)
  r_squared <- summary(lm_model)$r.squared
  return(list(correlation = correlation, r_squared = r_squared))
}

# Step 1: Filter standard samples
standards_df <- medianMFI_df %>%
  filter(str_detect(Sample, "Standard"))

# Step 2: Extract dilution factor
standards_df <- standards_df %>%
  mutate(Dilution_Factor = -as.numeric(str_extract(Sample, "\\d+")))

# Step 3: Prepare data for plotting
melted_df <- standards_df %>%
  select(-Sample) %>%
  pivot_longer(cols = all_of(valid_analytes), 
               names_to = "Analyte", 
               values_to = "MFI") %>%
  mutate(Log10_MFI = log10(MFI + 1))  # Adding 1 to avoid log(0)

# Step 4: Create standard curves for each analyte
plots_list <- lapply(unique(melted_df$Analyte), function(analyte) {
  # Filter data for the current analyte and remove NA/infinite values
  analyte_df <- filter(melted_df, Analyte == analyte & !is.na(Log10_MFI) & !is.infinite(Log10_MFI))
  
  # Check if analyte_df is empty after filtering
  if (nrow(analyte_df) == 0) {
    message(paste("No valid data for analyte:", analyte))
    return(NULL)
  }
  
  # Create plot
  plot <- ggplot(analyte_df, aes(x = Dilution_Factor, y = Log10_MFI)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") + # Add line of best fit
    labs(title = paste("Standard curve for", analyte),
         x = "Dilution Factor",
         y = "Log10(MFI)") +
    theme_minimal()
  
  # Calculate correlation and R-squared
  correlation_r_squared <- calculate_correlation(analyte_df)
  
  # Add R-squared value to the plot annotation
  plot <- plot + 
    annotate("text", x = Inf, y = -Inf, 
             label = paste("R^2 =", round(correlation_r_squared$r_squared, 2)), 
             hjust = 1, vjust = 0)
  
  # Append analyte and R-squared value to plate_qc data frame
  plate_qc <<- bind_rows(plate_qc, data.frame(Analyte = analyte, R_squared = correlation_r_squared$r_squared))
  
  return(plot)
})

# Combine all plots into a single object
combined_plots <- do.call(gridExtra::grid.arrange, c(plots_list, ncol = 2))

# Print combined plots
print(combined_plots)

# Print the plate_qc data frame with R-squared values
print(plate_qc)
```
```{r Plate QC- R squared calculation, echo=FALSE, warning=FALSE}
    # Extract the minimum R-squared value from the parameters
    min_Rsquared <- as.numeric(params$min_Rsquared)
    
    # Remove duplicate rows from plate_qc to ensure unique entries
    plate_qc <- plate_qc %>%
      distinct()
    
    # Create a column 'min_Rsq' indicating whether each analyte meets the minimum R-squared value
    plate_qc$min_Rsq <- plate_qc$R_squared > min_Rsquared
    plate_qc$min_Rsq <- as.logical(plate_qc$min_Rsq)
    
    # View the plate_qc data frame
    plate_qc
    
    # Check if any analyte meets the preset minimum R-squared value
    if (any(plate_qc$min_Rsq)) {
      # If at least one analyte meets the criteria
      print("This plate has passed plate-level quality control for standard curve linearity.")
      print("At least one analyte meets the criteria for the preset minimum R-squared (see min_Rsquared in YAML).")
      print("Median MFI from this plate will be included in the compiled study dataframe.")
    } else {
      # If no analytes meet the criteria
      print("This plate has failed plate-level QC for standard curves.")
      print("No analytes met the preset minimum R-squared (see min_Rsquared in YAML).")
      print("All MFI values from this plate will display as `Failed Plate QC`.")
    }
```

### Create quality controlled Median MFI data

```{r Plate QC- R squared check, echo=FALSE, warning=FALSE}
# Extract the minimum R-squared value from the parameters
min_Rsquared <- as.numeric(params$min_Rsquared)

# Remove duplicate rows from plate_qc to ensure unique entries
plate_qc <- plate_qc %>%
  distinct()

# Create a column 'min_Rsq' indicating whether each analyte meets the minimum R-squared value
plate_qc$min_Rsq <- plate_qc$R_squared > min_Rsquared
plate_qc$min_Rsq <- as.logical(plate_qc$min_Rsq)

# View the plate_qc data frame
plate_qc

# Check if any analyte meets the preset minimum R-squared value
if (any(plate_qc$min_Rsq)) {
  # If at least one analyte meets the criteria
  print("This plate has passed plate-level quality control for standard curve linearity.")
  print("At least one analyte meets the criteria for the preset minimum R-squared (see min_Rsquared in YAML).")
  print("Median MFI from this plate will be included in the compiled study dataframe.")
} else {
  # If no analytes meet the criteria
  print("This plate has failed plate-level QC for standard curves.")
  print("No analytes met the preset minimum R-squared (see min_Rsquared in YAML).")
  print("All MFI values from this plate will display as `Failed Plate QC`.")
}
```

```{r Plate QC- Final Clean Data, echo=FALSE, warning=FALSE}
#No input required
      # Define the output directory and filename
      output_dir <- "data/clean/luminex"
      output_file <- file.path(output_dir, paste0(tools::file_path_sans_ext(basename(params$input_file)), "_clean.csv"))
      
      # Create the output directory if it doesn't exist
      if (!dir.exists(output_dir)) {
        dir.create(output_dir, recursive = TRUE)
      }
      
      # Extract the plate name from the input file name
      plate_name <- tools::file_path_sans_ext(basename(params$input_file))
      
      # Identify the standards that met the minimum R-squared value
      passing_standards <- plate_qc$Analyte[plate_qc$min_Rsq]
      
      # Create a string of passing standards
      passing_standards_str <- paste(passing_standards, collapse = "; ")
      
      # Check if the plate passed the R-squared quality control
      if (any(plate_qc$min_Rsq)) {
        # If the plate passed QC, replace all NA values with "Failed QC (bead)"
        clean_medianMFI_df <- medianMFI_df %>%
          mutate(across(where(is.numeric), ~ ifelse(is.na(.), "Failed QC (bead)", .)))
        
        # Print a message indicating the plate passed QC
        cat("Plate passed QC. All NA values have been replaced with 'Failed QC (bead)'.\n")
      } else {
        # If the plate failed QC, set all median MFI data to "Failed QC (standards)"
        clean_medianMFI_df <- medianMFI_df %>%
          mutate(across(where(is.numeric), ~ "Failed QC (standards)"))
        
        # Print a message indicating the plate failed QC
        cat("Plate failed QC. All median MFI data set to 'Failed QC (standards)'.\n")
      }
      
      # Add the plate name and passing standards columns to the cleaned data
      clean_medianMFI_df <- clean_medianMFI_df %>%
        mutate(Plate = plate_name, 
               Standard = passing_standards_str)
      
      # Export the finalized data as a CSV file
      write.csv(clean_medianMFI_df, file = output_file, row.names = FALSE)
      
      # Print a message indicating the export was successful
      cat("The cleaned median MFI data has been exported to", output_file,". The plate name and standards which met the minimum R squared have been included in the cleaned median MFI file. \n")
```

### Developer notes

```{=html}
<!-- 
DEVELOPER NOTES
# Useful links BSA corrected: https://docs.google.com/spreadsheets/d/1kFEW7UqGGvfmOtYQq_AdP3V2w7LwdhDLpTHPmPaJGnM/edit#gid=1434210263

# TODO: Test with Cat 
```
-->

